---
title: "Tissue-disease-specific inflammatory gene signatures"
author: "chripla"
date: "`r Sys.Date()`"
output: html_document
params:
  database: "data/ibdome_v1.0.1.sqlite"
  artifact_dir: "results"
  de_results_dir: "results/DESeq2"
---

Load libraries:
```{r}
box::use(
   lib/colors[COLORS],
   lib/db[db_connect],
   dplyr[...],
   tidyr[pivot_wider],
   tibble[column_to_rownames],
   SummarizedExperiment[SummarizedExperiment, colData, assay, rowData, `rowData<-`],
   DESeq2[vst, plotPCA, DESeqDataSet, estimateSizeFactors],
   ggplot2[...],
   cowplot[theme_cowplot, plot_grid, background_grid],
   readr[write_csv, write_tsv]
)
``` 

Connect to the IBDome database:
```{r}
DB = db_connect(params$database)
dir.create(params$artifact_dir, showWarnings = FALSE)
dir.create(params$de_results_dir, showWarnings = FALSE )
``` 

### Get RNA-seq data from the database

```{r}
rnaseq_samples_all = tbl(DB, "samples_subjects_tissues_with_inflammation") |> 
  filter(sample_type == "RNA-seq") |>
  collect()
``` 

Filter for tissue coarse:
* ileum
* colon

disease:
* CD
* UC
* non-IBD

```{r}
rnaseq_samples = rnaseq_samples_all |>
  filter(disease %in% c("Crohn's disease", "non-IBD", "Ulcerative colitis"), tissue_coarse %in% c("colon", "ileum")) 
```

Get raw counts from database:
```{r}
count_mat = tbl(DB, "data_rnaseq") |> 
  filter(sample_id %in% rnaseq_samples$sample_id) |> 
  pivot_wider(id_cols = gene_id, names_from="sample_id", values_from="raw_counts") |>
  collect() 
```

Get histo-scores from database:
```{r}
histo_all = tbl(DB, "samples_subjects_tissues") |>
  inner_join(tbl(DB, "data_histopathology2"), by="sample_id") |> 
  select(sample_id, date, subject_id, normalized_naini_cortina_score, normalized_riley_score, tissue_coarse) |>
  collect()

histo_rnaseq = histo_all |> 
  inner_join(select(rnaseq_samples, rnaseq_sample_id=sample_id, subject_id, date, tissue_coarse), by=c("subject_id", "date", "tissue_coarse"), multiple="all") |> 
  group_by(rnaseq_sample_id) |>
  summarize(normalized_naini_cortina_score = median(normalized_naini_cortina_score, na.rm = TRUE),
            normalized_riley_score = median(normalized_riley_score, na.rm = TRUE)) 
```

Combine RNAseq samples with histo-scores:
```{r}
rnaseq_obs = rnaseq_samples |> 
  left_join(histo_rnaseq, by=c("sample_id"="rnaseq_sample_id")) 
``` 

To make samples comparable define a normalized inflammation score across Riley and Naini-Cortina scores:
```{r}
rnaseq_obs = rnaseq_obs |>
  mutate(normalized_inflammation_score = rowMeans(cbind(normalized_naini_cortina_score, normalized_riley_score), na.rm=TRUE))
``` 

## Inspect covariates with prinicipal component analysis

```{r}
adata = SummarizedExperiment(
  assays=list(counts=count_mat |> column_to_rownames("gene_id") |> as.matrix() |> round() |> (\(x) x[, rnaseq_obs$sample_id])()),
  colData=rnaseq_obs |> column_to_rownames("sample_id"),
  rowData = count_mat |> select(gene_id) |> left_join(collect(tbl(DB, "genes")), by=c("gene_id"="ensg"))
)
``` 

Define covariates to plot:
```{r}
covariates = c("disease", "tissue_coarse", "normalized_inflammation_score")
```

```{r, fig.height=3, fig.width=15}
dds = DESeqDataSet(adata, design=~1) 
dds = estimateSizeFactors(dds)
vsd = vst(dds)
pca_res = plotPCA(vsd, returnData=TRUE, intgroup="disease") 
pct_var = round(100 * attr(pca_res, "percentVar"))
pca_df = pca_res |> as_tibble() |> select(name, PC1, PC2) |> 
  inner_join(colData(dds) |> as_tibble(rownames="name"), by="name")
  
plots = lapply(covariates, function(x) {
    p = ggplot(pca_df, aes_string(x="PC1", y="PC2", color=x)) + 
      geom_point(size=2) + 
      theme_bw() +
      xlab(sprintf("PC1 (%s%% variance)", pct_var[1])) + 
      ylab(sprintf("PC2 (%s%% variance)", pct_var[2])) +
      theme(legend.position="right")
    
    # Check if the covariate is categorical or continuous
    if (x %in% names(COLORS)) {
      # Apply manual colors for categorical variables
      p = p + scale_color_manual(values=COLORS[[x]])
    } else if (is.numeric(pca_df[[x]])) {
      title_text = sub("_", "\n", x)  # Replace first underscore with a new line
      
      # Apply continuous color scale for numeric variables
      p = p + scale_color_viridis_c(option="magma") + 
        guides(color = guide_colorbar(title = title_text))
    }
    
    # Rename legend title from "tissue_coarse" to "tissue" only in this plot
    if (x == "tissue_coarse") {
      p = p + labs(color = "tissue")
    }
    # increase legend labels:
    p <- p + theme(legend.text = element_text(size = 12))
    # Force aspect ratio
    p <- p + theme(aspect.ratio = 1)

    # Convert ggplot object to a grob
    p_grob <- ggplotGrob(p)  # Use ggplotGrob instead of print

    return(p_grob) # Return the grob
 })

# Get the maximum width of the legend for consistent plot area
legend_widths <- sapply(plots, function(p) {
  legend_idx <- which(names(p$grobs) == "guide-box")
  if (length(legend_idx) > 0) {
    sum(p$grobs[[legend_idx]]$widths)
  } else {
    0 # No legend
  }
})
max_legend_width <- max(legend_widths)

# Add margin to each plot to account for the maximum legend width
plots <- lapply(plots, function(p) {
  # Convert grob back to ggplot object for adding margin
  p <- ggplotify::as.ggplot(p)
  p + theme(plot.margin = unit(c(5.5, max_legend_width, 5.5, 5.5), "pt"))
})

cowplot::plot_grid(plotlist = plots, ncol=3)
ggsave(file.path(params$artifact_dir, "Figure3a.pdf"), width = 15, height = 3)
``` 

## DE analysis


### Define inflammation-tissue groups
```{r}
rnaseq_sample_meta <- rnaseq_samples_all |>
  mutate(group_fine = case_when(disease=="non-IBD" ~ "nonIBD",
                                disease=="Crohn's disease" & inflammation_status == "inflamed" ~ "CD_inflamed",
                                disease=="Crohn's disease" & inflammation_status == "non_inflamed" ~ "CD_non_inflamed",
                                disease=="Ulcerative colitis" & inflammation_status == "inflamed" ~ "UC_inflamed",
                                disease=="Ulcerative colitis" & inflammation_status == "non_inflamed" ~ "UC_non_inflamed",
                                TRUE ~ "unknown"))

# add tissue groups and scale age:
rnaseq_sample_meta = rnaseq_sample_meta |>
    mutate(group_tissue = paste(group_fine, tissue_coarse, sep = "_")) |>
    mutate(age = as.numeric(format(as.Date(date), "%Y"))- birth_year) |>
    mutate(age_scaled = scale(age)[,1])
``` 

### Export tables to run DESeq2 with nextflow pipeline

### 1.) Meta data: 
```{r}
write_csv(rnaseq_sample_meta, file.path(params$de_results_dir, "meta_data_rnaseq.csv"))
``` 

### 2.) Counts table:
```{r}
# add gene names:
count_mat_all = tbl(DB, "data_rnaseq") |> 
  pivot_wider(id_cols = gene_id, names_from="sample_id", values_from="raw_counts") |>
  left_join(tbl(DB, "genes") |> select(ensg, hgnc_symbol) |> rename(gene_name=hgnc_symbol, gene_id = ensg)) |> 
  collect()

# correct order:
count_mat_all = count_mat_all[,c("gene_id", "gene_name", rnaseq_sample_meta$sample_id)]

# save as tsv:
write_tsv(count_mat_all, file.path(params$de_results_dir, "rnaseq_counts.tsv"))
``` 

### Session info

```{r}
sessionInfo()
``` 
