---
title: "Extract normalized histo scores for the existing images including disease information"
author: "sandrocarollo"
date: "`r Sys.Date()`"
output: html_document
params:
  database: "data/ibdome_v1.0.1.sqlite"
  resultdir: "./results/metadata_HE"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load libraries:

```{r}
box::use(
   dplyr[...],
   lib/db[db_connect],
)
```

Connect to the IBDome database:

```{r}
DB = db_connect(params$database)
dir.create(params$resultdir, showWarnings = FALSE)
```

### Getting Histology samples metadata

```{r}
samples_HE <- tbl(DB, "data_he_staining") |>
  left_join(tbl(DB, "samples")) |>
  select(c("sample_id", "orig_id", "ibdome_id", "subject_id", "date", "tissue", "sampling_procedure" )) |>
  collect()

samples_histo <- tbl(DB, "data_histopathology2") |>
  left_join(tbl(DB, "samples")) |>
  select(c("sample_id", "modified_naini_cortina_score", "modified_riley_score", "normalized_naini_cortina_score",  "normalized_riley_score", "ibdome_id", "date",  "tissue", "sampling_procedure")) |>
  collect()

samples_subjects_tissues <- tbl(DB, "samples_subjects_tissues") |>
  select("ibdome_id", "date", "tissue", "disease", "sex", "sampling_procedure") |>
  collect() |>
  unique()

samples_rna <- tbl(DB, "samples") |>
  filter(sample_type == "RNA-seq") |>
  select("subject_id", "ibdome_id", "date", "tissue", "sample_id") |>
  collect()

samples_all <- samples_HE |>
  left_join(samples_histo, by=c("ibdome_id", "date", "tissue", "sampling_procedure")) |>
  left_join(samples_subjects_tissues, by=c("ibdome_id", "date", "tissue", "sampling_procedure"))

# Getting Histo-RNA samples
samples_all_rna <- samples_all |>
  left_join(samples_rna, by = c("subject_id", "date", "tissue")) |>
  filter(!is.na(sample_id))
```

Getting Endoscopy score

```{r}
samples_Endo <- tbl(DB, "data_endoscopy") |>
  left_join(tbl(DB, "samples")) |>
  select(c("subject_id", "sample_id", "date", "SES_CD", "UCEIS" )) |>
  collect()
write.csv(samples_Endo, file.path(params$resultdir, "Endo_scores.csv"), row.names = FALSE, quote = FALSE)

```

Get duplicates:

```{r}
dups = samples_all[duplicated(samples_all$sample_id.x),]
dups_rna = samples_all_rna[duplicated(samples_all_rna$sample_id.x),]
```

Duplicates removal:

```{r}
# Group by the columns that should be unique and summarize the numeric columns
samples_all_unique <- samples_all %>%
  group_by(sample_id.x, subject_id, orig_id, ibdome_id, date, tissue, disease, sex, sampling_procedure) %>%
  summarize(
    modified_naini_cortina_score = mean(modified_naini_cortina_score, na.rm = TRUE),
    modified_riley_score = mean(modified_riley_score, na.rm = TRUE),
    normalized_naini_cortina_score = mean(normalized_naini_cortina_score, na.rm = TRUE),
    normalized_riley_score = mean(normalized_riley_score, na.rm = TRUE),
    .groups = 'drop'
  )

samples_all_unique_rna <- samples_all_rna %>%
  group_by(sample_id.x, subject_id, orig_id, date, tissue, disease, sex, sampling_procedure, sample_id) %>%
  summarize(
    modified_naini_cortina_score = mean(modified_naini_cortina_score, na.rm = TRUE),
    modified_riley_score = mean(modified_riley_score, na.rm = TRUE),
    normalized_naini_cortina_score = mean(normalized_naini_cortina_score, na.rm = TRUE),
    normalized_riley_score = mean(normalized_riley_score, na.rm = TRUE),
    .groups = 'drop'
  )

# Check if there are still duplicates
dups_after <- samples_all_unique[duplicated(samples_all_unique$sample_id.x),]
dups_rna_after <- samples_all_unique_rna[duplicated(samples_all_unique_rna$sample_id.x),]
```

Create the slide.csv input file for STAMP pipeline
NOTE: PATIENT refers only to sample in our case, not the actual patient.

```{r}
slide_data <- samples_all_unique %>%
  select(FILENAME = orig_id, PATIENT = sample_id.x)

# Add "_class_tokens" suffix to FILENAME column for Virchow2 model
slide_data_class_token <- slide_data %>%
  mutate(FILENAME = paste0(FILENAME, "_class_tokens"))

# Write slide_class_token.csv
write.csv(slide_data, file.path(params$resultdir, "slide.csv"), row.names = FALSE, quote = FALSE)
write.csv(slide_data_class_token, file.path(params$resultdir, "slide_class_token.csv"), row.names = FALSE, quote = FALSE)
```

Patients with at least one score

```{r}
# Rename sample_id.x to PATIENT
samples_all_unique <- samples_all_unique %>%
  rename(PATIENT = sample_id.x)
samples_all_unique_rna <- samples_all_unique_rna %>%
  rename(PATIENT = sample_id.x)

write.csv(samples_all_unique, file.path(params$resultdir, "samples_all_unique.csv"), row.names = FALSE, quote = FALSE)

# Generating Disease tables for classifier
# 1. Berlin+Erlangen
disease_all <- samples_all_unique %>%
  filter(disease %in% c("Crohn's disease", "Ulcerative colitis"))
write.csv(disease_all, file.path(params$resultdir, "clini_table_disease_all.csv"), row.names = FALSE, quote = FALSE)
# 2. Berlin subset
disease_berlin <- disease_all %>%
  filter(grepl("^ibdome_berlin", subject_id))
write.csv(disease_berlin, file.path(params$resultdir, "clini_table_disease_all_berlin.csv"), row.names = FALSE, quote = FALSE)
# 3. Erlangen subset
disease_erlangen <- disease_all %>%
  filter(grepl("^ibdome_erlangen", subject_id))
write.csv(disease_erlangen, file.path(params$resultdir, "clini_table_disease_all_erlangen.csv"), row.names = FALSE, quote = FALSE)
# 4. Erlangen inflamed subset
disease_erlangen_inflamed <- disease_erlangen %>%
  filter(
    (normalized_naini_cortina_score > 0) |
    (normalized_riley_score > 0)
  )
write.csv(disease_erlangen_inflamed, file.path(params$resultdir, "clini_table_disease_all_inflamed_erlangen.csv"), row.names = FALSE, quote = FALSE)

# Filter to include only patients with at least one score
samples_with_scores <- samples_all_unique %>%
  filter(!is.na(normalized_naini_cortina_score) | !is.na(normalized_riley_score))
samples_with_scores_rna <- samples_all_unique_rna %>%
  filter(!is.na(normalized_naini_cortina_score) | !is.na(normalized_riley_score))
# Verify the number of patients
cat("Number of patients with at least one score:", n_distinct(samples_with_scores$PATIENT), "\n")
```

Patients with Naini Cortina Score and Riley Score

```{r}
# Naini Cortina table
cortina_table <- samples_with_scores %>%
  filter(!is.na(normalized_naini_cortina_score)) %>%
  select(PATIENT, subject_id, orig_id, date, tissue, disease, sex, normalized_naini_cortina_score)
cortina_table_rna <- samples_with_scores_rna %>%
  filter(!is.na(normalized_naini_cortina_score)) %>%
  select(PATIENT, sample_id, subject_id, orig_id, date, tissue, disease, sex, normalized_naini_cortina_score)
# Riley table
riley_table <- samples_with_scores %>%
  filter(!is.na(normalized_riley_score)) %>%
  select(PATIENT, subject_id, orig_id, date, tissue, disease, sex, normalized_riley_score)
riley_table_rna <- samples_with_scores_rna %>%
  filter(!is.na(normalized_riley_score)) %>%
  select(PATIENT, sample_id, subject_id, orig_id, date, tissue, disease, sex, normalized_riley_score)
# Verify the number of patients in each table
cat("Number of patients in Cortina table:", n_distinct(cortina_table$PATIENT), "\n")
cat("Number of patients in Riley table:", n_distinct(riley_table$PATIENT), "\n")
```

Removing patients scored with the wrong scoring system

```{r}
# Remove UC patients from Naini Cortina table
cortina_table_cleaned <- cortina_table %>%
  filter(disease != "Ulcerative colitis")
cortina_table_rna_cleaned <- cortina_table_rna %>%
  filter(disease != "Ulcerative colitis")
# Remove CD patients from Riley table
riley_table_cleaned <- riley_table %>%
  filter(disease != "Crohn's disease")
riley_table_rna_cleaned <- riley_table_rna %>%
  filter(disease != "Crohn's disease")
# Verify the number of patients in each table after cleaning
cat("Number of patients in Cortina table after cleaning:", n_distinct(cortina_table_cleaned$PATIENT), "\n")
cat("Number of patients in Riley table after cleaning:", n_distinct(riley_table_cleaned$PATIENT), "\n")
```

Splitting the Naini Cortina and Riley table for Berlin and Erlangen cohorts

```{r}
# Split Cortina table into Berlin and Erlangen
cortina_berlin <- cortina_table_cleaned %>%
  filter(grepl("^TRR241-B-", orig_id)) %>%
  select(-orig_id)  # Exclude orig_id
cortina_berlin_rna <- cortina_table_rna_cleaned %>%
  filter(grepl("^TRR241-B-", orig_id)) %>%
  select(-orig_id)  # Exclude orig_id

cortina_erlangen <- cortina_table_cleaned %>%
  filter(grepl("^TRR241-E-", orig_id)) %>%
  select(-orig_id)  # Exclude orig_id
cortina_erlangen_rna <- cortina_table_rna_cleaned %>%
  filter(grepl("^TRR241-E-", orig_id)) %>%
  select(-orig_id)  # Exclude orig_id

# Split Riley table into Berlin and Erlangen
riley_berlin <- riley_table_cleaned %>%
  filter(grepl("^TRR241-B-", orig_id)) %>%
  select(-orig_id)  # Exclude orig_id
riley_berlin_rna <- riley_table_rna_cleaned %>%
  filter(grepl("^TRR241-B-", orig_id)) %>%
  select(-orig_id)  # Exclude orig_id

riley_erlangen <- riley_table_cleaned %>%
  filter(grepl("^TRR241-E-", orig_id)) %>%
  select(-orig_id)  # Exclude orig_id
riley_erlangen_rna <- riley_table_rna_cleaned %>%
  filter(grepl("^TRR241-E-", orig_id)) %>%
  select(-orig_id)  # Exclude orig_id

# Verify the number of patients in each table
cat("Cortina Berlin entries:", nrow(cortina_berlin), "\n")
cat("Cortina Erlangen entries:", nrow(cortina_erlangen), "\n")
cat("Riley Berlin entries:", nrow(riley_berlin), "\n")
cat("Riley Erlangen entries:", nrow(riley_erlangen), "\n")
```

Saving all the tables

```{r}
# Save Naini Cortina tables
write.csv(cortina_table_cleaned, file.path(params$resultdir, "clini_table_cortina.csv"), row.names = FALSE, quote = FALSE)
write.csv(cortina_berlin, file.path(params$resultdir, "clini_table_cortina_berlin.csv"), row.names = FALSE, quote = FALSE)
write.csv(cortina_erlangen, file.path(params$resultdir, "clini_table_cortina_erlangen.csv"), row.names = FALSE, quote = FALSE)
write.csv(cortina_table_rna_cleaned, file.path(params$resultdir, "clini_table_cortina_rna.csv"), row.names = FALSE, quote = FALSE)
write.csv(cortina_berlin_rna, file.path(params$resultdir, "clini_table_cortina_berlin_rna.csv"), row.names = FALSE, quote = FALSE)
write.csv(cortina_erlangen_rna, file.path(params$resultdir, "clini_table_cortina_erlangen_rna.csv"), row.names = FALSE, quote = FALSE)
# Save Riley tables
write.csv(riley_table_cleaned, file.path(params$resultdir, "clini_table_riley.csv"), row.names = FALSE, quote = FALSE)
write.csv(riley_berlin, file.path(params$resultdir, "clini_table_riley_berlin.csv"), row.names = FALSE, quote = FALSE)
write.csv(riley_erlangen, file.path(params$resultdir, "clini_table_riley_erlangen.csv"), row.names = FALSE, quote = FALSE)
write.csv(riley_table_rna_cleaned, file.path(params$resultdir, "clini_table_riley_rna.csv"), row.names = FALSE, quote = FALSE)
write.csv(riley_berlin_rna, file.path(params$resultdir, "clini_table_riley_berlin_rna.csv"), row.names = FALSE, quote = FALSE)
write.csv(riley_erlangen_rna, file.path(params$resultdir, "clini_table_riley_erlangen_rna.csv"), row.names = FALSE, quote = FALSE)
```
